Bayesian inference of relative fitness on high-throughput pooled competition assays

### 2023-10-11 doc/_abstract.qmd
add missing sections to paper

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{lineno}
\linenumbers
\usepackage[font={footnotesize, sf}, labelfont=bf, margin=0.05\linewidth]{caption}
\renewcommand{\familydefault}{\sfdefault}
% Bold the 'Figure #' in the caption and separate it from the title/caption
% with a period
% Captions will be left justified
\usepackage[
	aboveskip=1pt,
	labelfont=bf,
	labelsep=period,
	justification=raggedright,
	singlelinecheck=off
]{caption}

% Add numbered lines
\usepackage{lineno}
\linenumbers

% Package to include multiple title pages
% This will allow me to add a tile to the main text and to the SI
\usepackage{titling}

% Define command to begin the supplementary section
\newcommand{\beginsupplement}{
\setcounter{section}{0} % Restart section counter
\renewcommand{\thesection}{S\arabic{section}}%
\setcounter{table}{0} % Restart table counter
\renewcommand{\thetable}{S\arabic{table}}%
\setcounter{figure}{0} % Restart figure counter
\renewcommand{\thefigure}{S\arabic{figure}}%
\setcounter{equation}{0} % Restart equation counter
\renewcommand{\theequation}{S\arabic{equation}}%
}

% Add author affiliations
\usepackage{authblk}

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

% References
% \usepackage[]{biblatex}
\usepackage[
	backend=bibtex,
	style=phys, 
	sorting=none,				% Do not sort bibliography
	url=false, 					% Do not show url in reference
	doi=false, 					% Do not show doi in reference
	isbn=false, 				% Do not show isbn link in reference
	eprint=false, 			% Do not show eprint link in reference
	maxbibnames=9, 			% Include up to 9 names in citation
	firstinits=true,
]{biblatex}
\addbibresource{references.bib}

\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdfauthor={Manuel Razo-Mejia; Madhav Mani; Dmitri Petrov},
  pdfkeywords={bayesian inference, variational inference, relative
fitness, open-source, reproducible research},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

% Title
\title{Bayesian inference of relative fitness on high-throughput pooled competition assays}

% Authors
\author[1]{Manuel Razo-Mejia}
\author[2,3]{Madhav Mani}
\author[1,4,*]{Dmitri Petrov}

% Affiliations
\affil[1]{Department of Biology, Stanford University, Stanford, CA, USA}
\affil[2]{Department of Engineering Sciences and Applied Mathematics, Northwestern University, Chicago, IL, USA}
\affil[3]{NSF-Simons Center for Quantitative Biology, Northwestern University, Chicago, IL, USA}
\affil[4]{Stanford Cancer Institute, Stanford University School of Medicine, Stanford, CA, USA}
\affil[*]{Correspondence: dpetrov@stanford.edu}

\setcounter{Maxaffil}{0}
% Set affiliations in small font
\renewcommand\Affilfont{\itshape\small}

\date{}

\begin{document}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, interior hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

% Remove main text from the table of contents by specifying not to include
% any section or subsection
\addtocontents{toc}{\protect\setcounter{tocdepth}{-1}}

% Define reference segment for main text
\begin{refsegment}
% Generate filter to not include references from main text in the
% supplemental references
\defbibfilter{notother}{not segment=\therefsegment}

\maketitle

\begin{abstract}
    The tracking of lineage frequencies via DNA barcode sequencing enables the
  quantification of microbial fitness. However, experimental noise coming from
  biotic and abiotic sources complicates the computation of a reliable
  inference. We present a Bayesian pipeline to infer relative microbial fitness
  from high-throughput lineage tracking assays. Our model accounts for multiple
  sources of noise and propagates uncertainties throughout all parameters in a
  systematic way. Furthermore, using modern variational inference methods based
  on automatic differentiation, we are able to scale the inference to a large
  number of unique barcodes. We extend this core model to analyze
  multi-environment assays, replicate experiments, and barcodes linked to
  genotypes. On simulations, our method recovers known parameters within
  posterior credible intervals. This work provides a generalizable Bayesian
  framework to analyze lineage tracking experiments. The accompanying
  open-source software library enables the adoption of principled statistical
  methods in experimental evolution.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The advent of DNA barcoding---the ability to uniquely identify cell
lineages with DNA sequences integrated at a specific locus---and
high-throughput sequencing has opened new venues for understanding
microbial evolutionary dynamics with an unprecedented level of temporal
resolution \autocite{levy2015,nguyenba2019a,ascensao2023}. These
experimental efforts rely on our ability to reliably infer the relative
fitness of an ensemble of diverse genotypes. Moreover, inferring these
fitness values over an ensemble of environmental conditions can help us
determine the phenotypic diversity of a rapid adaptation process
\autocite{kinsler2020}.

As with any other sequencing-based quantification, tracking lineages via
DNA barcode sequencing is inexorably accompanied by noise sources coming
from experimental manipulation of the microbial cultures, DNA
extraction, and sequencing library preparation that involves multiple
rounds of PCR amplification, and the sequencing process itself. Thus,
accounting for the uncertainty when inferring the relevant parameters
from the data is a crucial step to draw reliable conclusions. Bayesian
statistics presents a paradigm by which one can account for all known
sources of uncertainty in a principled way \autocite{eddy2004a}. This,
combined with the development of modern Markov Chain Monte Carlo
sampling algorithms \autocite{betancourt2017} and approximate
variational approaches \autocite{kucukelbir2016} have boosted a
resurgence of Bayesian methods in different fields
\autocite{efron2013a}.

We present a Bayesian inference pipeline to quantify the uncertainty
about the parametric information we can extract from high-throughput
competitive fitness assays given a model of the data generation process
and experimental data. In these assays, the fitness of an ensemble of
genotypes is determined relative to a reference genotype
\autocite{kinsler2020,ascensao2023}. Figure~\ref{fig-01}(A) shows a
schematic of the experimental procedure in which an initial pool of
barcoded strains are mixed with a reference strain and inoculated into
fresh media. After some time---usually, enough time for the culture to
saturate---an aliquot is transferred to fresh media, while the remaining
culture is used for DNA sequencing of the lineage barcodes. The
time-series information of the relative abundance of each lineage, i.e.,
the barcode frequency depicted in Figure~\ref{fig-01}(B), is used to
infer the relative fitness---the growth advantage on a per-cycle
basis---for each lineage with respect to the reference strain. The
proposed statistical model accounts for multiple sources of uncertainty
when inferring the lineages' relative fitness values (see
Section~\ref{sec-experiment} for details on sources of uncertainty
accounted for by the model). Furthermore, minor changes to the core
statistical model allow us to account for relevant experimental
variations of these competition assays. More specifically, in
Section~\ref{sec-multienv}, we present a variation of the statistical
model to infer fitness on growth dilution cycles in multiple
environments with proper error propagation. Furthermore, as described in
Section~\ref{sec-replicates}, our statistical model can account for
batch-to-batch differences when jointly analyzing multiple experimental
replicates using a Bayesian hierarchical model. Finally, a variant of
these hierarchical models, presented in Section~\ref{sec-genotypes}, can
account for variability within multiple barcodes mapping to equivalent
genotypes within the same experiment.

For all the model variations presented in this paper, we benchmark the
ability of our pipeline to infer relative fitness parameters against
synthetic data generated from logistic growth simulations with added
random noise. A \texttt{Julia} package accompanies the present method to
readily implement the inference pipeline with state-of-the-art
scientific computing software.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{sec-experiment}{%
\subsection{Experimental setup}\label{sec-experiment}}

The present work is designed to analyze time-series data of relative
abundance of multiple microbial lineages uniquely identified by a DNA
barcode \autocite{kinsler2020,ascensao2023}. In these competition
assays, an ensemble of genotypes is pooled together with an unlabeled
reference strain that, initially, represents the vast majority
(\(\geq 90\%\)) of the cells in the culture (see schematic in
Figure~\ref{fig-01}(A)). Furthermore, a fraction of labeled genotypes
equivalent to the unlabeled reference strain---hereafter defined as
\emph{neutral} lineages---are spiked in at a relatively high abundance
(\(\approx 3-5\%\)). The rest of the culture is left for the ensemble of
genotypes of interest.

To determine the relative fitness of the ensemble of genotypes, a series
of growth-dilution cycles are performed on either a single or multiple
environments. In other words, the cultures are grown for some time;
then, an aliquot is inoculated into fresh media for the next growth
cycle. This process is repeated for roughly 4-7 cycles, depending on the
initial abundances of the mutants and their relative growth rates. The
DNA barcodes are sequenced at the end of each growth cycle to quantify
the relative abundance of each of the barcodes. We point the reader to
\textcite{kinsler2020} for specific details on these assays for \emph{S.
cerevisiae} and to \textcite{ascensao2023} for equivalent assays for
\emph{E. coli}. Figure~\ref{fig-01}(B) presents a typical barcode
trajectory where the black trajectories represent the so-called
\emph{neutral lineages}, genetically equivalent to the untagged ancestor
strain that initially dominates the culture. These spiked-in neutral
lineages simplify the inference problem since the fitness metric of all
relevant barcodes is quantified with respect to these barcodes---thus
referred to as \emph{relative fitness}.

\begin{figure}

{\centering \includegraphics{./figs/fig01.pdf}

}

\caption{\label{fig-01}\textbf{Typical competitive fitness experiment}.
(A) Schematic of the typical experimental design to determine the
competitive fitness of an ensemble of barcoded genotypes. Genotypes are
pooled together and grown over multiple growth-dilution cycles. At the
end of each cycle, a sample is processed to generate a library for
amplicon sequencing. (B) Typical barcode trajectory dataset. From each
time point, the relative frequency of each barcode is determined from
the total number of reads. Shades of blue represent different relative
fitness. Darker gray lines define the typical trajectory of neutral
lineages.}

\end{figure}

\hypertarget{preliminaries-on-mathematical-notation}{%
\subsection{Preliminaries on mathematical
notation}\label{preliminaries-on-mathematical-notation}}

Before jumping directly into the Bayesian inference pipeline, let us
establish the mathematical notation used throughout this paper. We
define (column) vectors as underlined lowercase symbols such as
\begin{equation}\protect\hypertarget{eq-vec_def}{}{
\underline{x} = \begin{bmatrix}
    x_1\\
    x_2\\
    \vdots\\
    x_N
\end{bmatrix}.
}\label{eq-vec_def}\end{equation} In the same way, we define matrices as
double-underline uppercase symbols such as
\begin{equation}\protect\hypertarget{eq-mat_def}{}{
\underline{\underline{A}} =
\begin{bmatrix}
    A_{11} & A_{12} & \cdots & A_{1N}\\
    A_{21} & A_{22} & \cdots & A_{2N}\\
    \vdots & \vdots & \ddots & \vdots\\
    A_{M1} & A_{M2} & \cdots & A_{MN}\\
\end{bmatrix}.
}\label{eq-mat_def}\end{equation}

\hypertarget{sec-fitness_model}{%
\subsection{Fitness model}\label{sec-fitness_model}}

Empirically, each barcode frequency trajectory follows an exponential
function of the form \autocite{levy2015,kinsler2020,ascensao2023}
\begin{equation}\protect\hypertarget{eq-fitness}{}{
f_{t+1}^{(b)} = f_{t}^{(b)} \mathrm{e}^{(s^{(b)} - \bar{s}_t)\tau},
}\label{eq-fitness}\end{equation} where \(f_{t}^{(b)}\) is the frequency
of barcode \(b\) at the end of cycle number \(t\), \(s^{(b)}\) is the
relative fitness with respect to the reference strain---the quantity we
want to infer from the data---\(\bar{s}_t\) is the mean fitness of the
culture at the end of cycle number \(t\), and \(\tau\) is the time pass
between cycle \(t\) and \(t+1\). We can rewrite
Equation~\ref{eq-fitness} as
\begin{equation}\protect\hypertarget{eq-logfreq}{}{
\frac{1}{\tau}\ln \frac{f_{t+1}^{(b)}}{f_{t}^{(b)}} = (s^{(b)} - \bar{s}_t).
}\label{eq-logfreq}\end{equation} Equation~\ref{eq-logfreq} separates
the measurements---the barcode frequencies---from the unobserved
(sometimes referred to as latent) parameters we want to infer from the
data---the population mean fitness and the barcode relative fitness.
This is ultimately the functional form used in our inference pipeline.
Therefore, the relative fitness is computed by knowing the log frequency
ratio of each barcode throughout the growth-dilution cycles.

The presence of the neutral lineages facilitates the determination of
the population mean fitness value \(\bar{s}_t\). Since every relative
fitness is determined relative to the neutral lineage that dominates the
culture, we define their fitness to be \(s^{(n)} = 0\), where the
superscript \((n)\) specifies their neutrality. This means that
Equation~\ref{eq-logfreq} for a neutral lineage takes the simpler form
\begin{equation}\protect\hypertarget{eq-logfreq_neutral}{}{
\frac{1}{\tau}\ln \frac{f_{t+1}^{(n)}}{f_{t}^{(n)}} = - \bar{s}_t.
}\label{eq-logfreq_neutral}\end{equation} Therefore, we can use the data
from these reference barcodes to directly infer the value of the
population mean fitness.

It is important to notice that the frequencies \(f_{t}^{(b)}\) are not
the allele frequencies in the population (most of the culture is not
sequenced since the reference strain is not barcoded), but rather the
relative frequencies in the total number of sequencing reads. A way to
conceptualize this subtle but important point is to assume exponential
growth in the \emph{number of cells} \(N_t^{(b)}\) of the form
\begin{equation}\protect\hypertarget{eq-ncells}{}{
N_{t+1}^{(b)} = N_{t}^{(b)} \mathrm{e}^{\lambda^{(b)}\tau},
}\label{eq-ncells}\end{equation} for every barcode \(b\) with growth
rate \(\lambda^{(b)}\). However, when we sequence barcodes, we do not
directly measure the number of cells, but some number of reads
\(r_t^{(b)}\) that map to barcode \(b\). In the simplest possible
scenario, we assume \begin{equation}\protect\hypertarget{eq-r_to_n}{}{
r_{t}^{(b)} \propto N_{t}^{(b)},
}\label{eq-r_to_n}\end{equation} where, importantly, the proportionality
constant depends on the total number of reads for the library for cycle
\(t\), which might vary from library to library. Therefore, to compare
the number of reads between libraries at different time points, we must
normalize the number of reads to the same scale. The simplest form is to
define a relative abundance, i.e., a frequency with respect to the total
number of reads, \begin{equation}\protect\hypertarget{eq-r_to_n_norm}{}{
f_{t}^{(b)} \equiv \frac{r_{t}^{(b)}}{\sum_{b'} r_{t}^{(b')}}.
}\label{eq-r_to_n_norm}\end{equation} This is the frequency
Equation~\ref{eq-fitness} describes.

Our ultimate objective is to infer the relative fitness \(s^{(b)}\) for
each of the \(M\) relevant barcodes in the experiment---hereafter
referred to as \(s^{(m)}\) to distinguish from the general \(s^{(b)}\)
and the neutral lineages \(s^{(n)}\) relative fitness. To do so, we
account for the three primary sources of uncertainty in our model:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Uncertainty in the determination of frequencies. Our model relates
  frequencies between adjacent growth-dilution cycles to the fitness of
  the corresponding strain. However, we do not directly measure
  frequencies. Instead, our data for each barcode consists of a length
  \(T\) vector of counts \(\underline{r}^{(b)}\) for each of the \(T\)
  cycles in which the measurements were taken.
\item
  Uncertainty in the value of the population mean fitness. We define
  neutral lineages to have fitness \(s^{(n)} = 0\), helping us anchor
  the value of the population mean fitness \(\bar{s}_t\) for each pair
  of adjacent growth cycles. Moreover, we take this parameter as an
  empirical parameter to be obtained from the data, meaning that we do
  not impose a functional form that relates \(\bar{s}_t\) to
  \(\bar{s}_{t+1}\). Thus, we must infer the \(T-1\) values of this
  population mean fitness with their uncertainty that must be propagated
  to the value of the mutants' relative fitness.
\item
  Uncertainty in each of the mutants' fitness values.
\end{enumerate}

To account for all these sources of uncertainty in a principled way, in
the next section, we develop a Bayesian inference pipeline.

\hypertarget{sec-bayesian_inference}{%
\subsection{Bayesian inference}\label{sec-bayesian_inference}}

As defined in Section~\ref{sec-fitness_model}, our ultimate objective is
to infer the vector of relative fitness values
\begin{equation}\protect\hypertarget{eq-fitness_mut_vec}{}{
\underline{s}^M = (s^{(1)}, s^{(2)}, \ldots, s^{(M)})^\dagger,
}\label{eq-fitness_mut_vec}\end{equation} where \(^\dagger\) indicates
the transpose. Our data consists of an \(T \times B\) matrix
\(\underline{\underline{R}}\), where \(B = M + N\) is the number of
unique barcodes given by the sum of the number of unique, relevant
barcodes we care about, \(M\), and the number of unique neutral
barcodes, \(N\), and \(T\) is the number of growth cycles where
measurements were taken. The data matrix is then of the form
\begin{equation}\protect\hypertarget{eq-R_mat}{}{
\underline{\underline{R}} = \begin{bmatrix}
- & \underline{r}_1 & - \\
- & \underline{r}_2 & - \\
 & \vdots & \\
- & \underline{r}_T & - \\
\end{bmatrix},
}\label{eq-R_mat}\end{equation} where each row \(\underline{r}_t\) is a
\(B\)-dimensional array containing the raw barcode counts at cycle
\(t\). We can further split each vector \(\underline{r}_t\) into two
vectors of the form
\begin{equation}\protect\hypertarget{eq-r-vec_split}{}{
\underline{r}_t = \begin{bmatrix}
\underline{r}_t^{N} \\
\underline{r}_t^{M}
\end{bmatrix},
}\label{eq-r-vec_split}\end{equation} i.e., the vector containing the
neutral lineage barcode counts \(\underline{r}_t^{N}\) and the
corresponding vector containing the mutant barcode counts
\(\underline{r}_t^{M}\). Following the same logic, matrix
\(\underline{\underline{R}}\) can be split into two matrices as
\begin{equation}\protect\hypertarget{eq-R-mat_split}{}{
\underline{\underline{R}} = \left[ 
\underline{\underline{R}}^N \; \underline{\underline{R}}^M
\right],
}\label{eq-R-mat_split}\end{equation} where
\(\underline{\underline{R}}^N\) is a \(T \times N\) matrix with the
barcode reads time series for each neutral lineage and
\(\underline{\underline{R}}^M\) is the equivalent \(T \times M\) matrix
for the non-neutral lineages.

Our objective is to compute the joint probability distribution for all
relative fitness values given our data. We can express this joint
posterior distribution using Bayes theorem as
\begin{equation}\protect\hypertarget{eq-bayes_obj}{}{
\pi(\underline{s}^M \mid \underline{\underline{R}}) = \frac{
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M)}
{\pi(\underline{\underline{R}})},
}\label{eq-bayes_obj}\end{equation} where hereafter \(\pi(\cdot)\)
defines a probability density function. When defining our statistical
model, we need not to focus on the denominator on the right-hand side of
Equation~\ref{eq-bayes_obj}. Thus, we can write
\begin{equation}\protect\hypertarget{eq-bayes_obj_propto}{}{
\pi(\underline{s}^M \mid \underline{\underline{R}}) \propto
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M).
}\label{eq-bayes_obj_propto}\end{equation} However, when implementing
the model computationally, the normalization constant on the right-hand
side of Equation~\ref{eq-bayes_obj} must be computed. This can be done
from the definition of the model via an integral of the form
\begin{equation}\protect\hypertarget{eq-bayes_evidence}{}{
\pi(\underline{\underline{R}}) = \int d^M \underline{s}^M
\pi(\underline{\underline{R}} \mid \underline{s}^M) 
\pi(\underline{s}^M),
}\label{eq-bayes_evidence}\end{equation} also known as a marginalization
integral. Hereafter, differentials of the form \(d^n\) imply a
\(n\)-dimensional integral.

Although Equation~\ref{eq-bayes_obj} and
Equation~\ref{eq-bayes_obj_propto} seem simple enough, recall that
Equation~\ref{eq-fitness} relates barcode frequency values and the
population mean fitness to the mutant relative fitness. Therefore, we
must include these nuisance parameters as part of our inference problem.
We direct the reader to the supplementary materials for the exact
definitions of these parameters. Here, it suffices to say that the
inference problem must include the vector \(\underline{\bar{s}}_T\) of
all population mean fitness values and the matrix
\(\underline{\underline{F}}\) of all barcode frequencies within the
sequencing data. With these nuisance variables in hand, the full
inference problem we must solve takes the form
\begin{equation}\protect\hypertarget{eq-bayes_full}{}{
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) \propto
\pi(
    \underline{\underline{R}} \mid
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
)
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}}
).
}\label{eq-bayes_full}\end{equation} To recover the marginal
distribution over the non-neutral barcodes relative fitness values, we
can numerically integrate out all nuisance parameters, i.e.,
\begin{equation}\protect\hypertarget{eq-bayes_full_marginal}{}{
\pi(\underline{s}^M \mid \underline{\underline{R}}) =
\int d^{T-1}\underline{\bar{s}}_T
\int d^{B}\underline{f}_1 \cdots
\int d^{B}\underline{f}_T
\;
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
).
}\label{eq-bayes_full_marginal}\end{equation}

\hypertarget{seq-split_posterior}{%
\subsubsection{Factorizing the posterior
distribution}\label{seq-split_posterior}}

The left-hand side of Equation~\ref{eq-bayes_full} is extremely
difficult to work with. However, we can take advantage of the structure
of our inference problem to rewrite it in a more manageable form.
Specifically, the statistical dependencies of our observations and
latent variables allow us to factorize the joint distribution into the
product of multiple conditional distributions. To gain some intuition
about this factorization, let us focus on the inference of the
population mean fitness values \(\underline{\bar{s}}_T\).
Equation~\ref{eq-logfreq_neutral} relates the value of the population
mean fitness to the neutral lineage frequencies and nothing else. This
suggests that when writing the posterior for these population mean
fitness parameters, we should be able to condition it only on the
neutral lineage frequency values, i.e.,
\(\pi(\underline{\bar{s}}_T \mid \underline{\underline{F}}^N)\). We
point the reader to Section~\ref{sec-bayes_def} for the full
mathematical details on this factorization. For our purpose here, it
suffices to say we can rewrite the joint probability distribution as a
product of conditional distributions of the form
\begin{equation}\protect\hypertarget{eq-split_posterior}{}{
\pi(
    \underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}} \mid
    \underline{\underline{R}}
) =
\pi(
    \underline{s}^M \mid \underline{\bar{s}}_T, \underline{\underline{F}}^M
)
\pi(
    \underline{\bar{s}}_T \mid \underline{\underline{F}}^N
)
\pi(\underline{\underline{F}} \mid \underline{\underline{R}}).
}\label{eq-split_posterior}\end{equation} Written in this form,
Equation~\ref{eq-split_posterior} captures the three sources of
uncertainty listed in Section~\ref{sec-fitness_model} in each term.
Starting from right to left, the first term on the right-hand side of
Equation~\ref{eq-split_posterior} accounts for the uncertainty when
inferring the frequency values given the barcode reads. The second term
accounts for the uncertainty in the values of the mean population
fitness at different time points. The last term accounts for the
uncertainty in the parameter we care about---the mutants' relative
fitnesses. We refer the reader to Section~\ref{sec-bayes_def} for an
extended description of the model with specific functional forms for
each term on the left-hand side of Equation~\ref{eq-split_posterior} as
well as the extension of the model to account for multiple experimental
replicates or hierarchical genotypes.

\hypertarget{variational-inference}{%
\subsubsection{Variational Inference}\label{variational-inference}}

One of the technical challenges to the adoption of Bayesian methods is
the analytical intractability of integrals such as that of
Equation~\ref{eq-bayes_full_marginal}. Furthermore, even though
efficient Markov Chain Monte Carlo (MCMC) algorithms such as Hamiltonian
Montecarlo can numerically perform this integration
\autocite{betancourt2017}, the dimensionality of the problem in
Equation~\ref{eq-split_posterior} makes an MCMC-based approach
prohibitively slow.

To overcome this computational limitation, we rely on the recent
development of the automatic differentiation variational inference
algorithm (ADVI) \autocite{kucukelbir2016}. Briefly, when performing
ADVI, our target posterior distribution
\(\pi(\theta \mid \underline{\underline{R}})\), where
\(\theta = (\underline{s}^M, \underline{\bar{s}}_T, \underline{\underline{F}})\),
is replaced by an approximate posterior distribution \(q_\phi(\theta)\),
where \(\phi\) fully parametrizes the approximate distribution. As
further explained in Section~\ref{sec-vi_primer}, the numerical
integration problem is replaced by an optimization problem of the form
\begin{equation}\protect\hypertarget{eq-vi_objective}{}{
q^*_\phi(\theta) = \min _\phi
D_{KL}(
    q_\phi(\theta) \lvert \lvert
    \pi(\theta \mid \underline{\underline{R}})
),
}\label{eq-vi_objective}\end{equation} where \(D_{KL}\) is the
Kulback-Leibler divergence. In other words, the complicated
high-dimensional numerical integration problem is transformed into a
much simpler problem of finding the value of the parameters \(\phi\)
such that Equation~\ref{eq-vi_objective} is satisfied as best as
possible within some finite computation time. Although to compute
Equation~\ref{eq-vi_objective}, we require the posterior distribution we
are trying to approximate
\(\pi(\theta \mid \underline{\underline{R}})\), it can be shown that
maximizing the so-called evidence lower bound (ELBO)
\autocite{kingma2014}---equivalent to minimizing the variational free
energy \autocite{gottwald2020}---is mathematically equivalent to
performing the optimization prescribed by
Equation~\ref{eq-vi_objective}. We direct the reader to
Section~\ref{sec-vi_primer} for a short primer on variational inference.

This work is accompanied by the Julia library \texttt{BarBay.jl} that
makes use of the implementation of both MCMC-based integration as well
as ADVI optimization to numerically approximate the solution of
Equation~\ref{eq-bayes_full_marginal} within the Julia ecosystem
\autocite{ge2018}.

\hypertarget{inference-on-a-single-dataset}{%
\subsection{Inference on a single
dataset}\label{inference-on-a-single-dataset}}

To assess the inference pipeline performance, we applied it to a
simulated dataset with known ground truth relative fitness values (See
Section~\ref{sec-logistic} for details on simulation).
Figure~\ref{fig-02}(A) shows the structure of the synthetic dataset. The
majority of barcodes of interest (faint color lines) are adaptive
compared to the neutral barcodes (\(s^{(m)} > 0\)). Although the barcode
frequency trajectories look relatively smooth, our fitness model
requires the computation of the log frequency ratio between adjacent
time points as derived in Equation~\ref{eq-logfreq}.
Figure~\ref{fig-02}(B) shows such data transformation where we can
better appreciate the observational noise input into our statistical
model. This noise is evident for the darker lines representing the
neutral barcodes since all of these lineages are assumed to be
identically distributed.

To visualize the performance of our inference pipeline in fitting our
fitness model to the observed data, we compute the so-called posterior
predictive checks (PPC). In short, the PPC consists of repeatedly
generating synthetic datasets in agreement with the results from the
inference results. In other words, we use the resulting parameter values
from the ADVI inference to generate possible datasets in agreement with
the inferred values (See Section~\ref{sec-ppc} for further details on
these computations). Figure~\ref{fig-02}(C) shows these results for all
neutral lineages (upper left corner plot) and a few representative
non-neutral barcodes. The different color shades represent the 95\%,
68\%, and 5\% credible regions, i.e., the regions where we expect to
find the data with the corresponding probability---or in terms of our
parameter, the \(X\%\) credible region is the interval where we expect
the true parameter value to lie with \(X\%\) probability.

>>>>

The tracking of lineage frequencies via DNA barcode sequencing enables the
quantification of microbial fitness. However, experimental noise coming from
biotic and abiotic sources complicates the computation of a reliable inference.
We present a Bayesian pipeline to infer relative microbial fitness from
high-throughput lineage tracking assays. Our model accounts for multiple sources
of noise and propagates uncertainties throughout all parameters in a systematic
way. Furthermore, using modern variational inference methods based on automatic
differentiation, we are able to scale the inference to a large number of unique
barcodes. We extend this core model to analyze multi-environment assays,
replicate experiments, and barcodes linked to genotypes. On simulations, our
method recovers known parameters within posterior credible intervals. This work
provides a generalizable Bayesian framework to analyze lineage tracking
experiments. The accompanying open-source software library enables the adoption
of principled statistical methods in experimental evolution.
